---
title: "Accesing Clean UNAIDS Data"
output: rmarkdown::html_vignette
description: |
  This vignette addresses how to pull the clean 2020 UNAIDS Estimates from Google Drive to R.
vignette: >
  %\VignetteIndexEntry{accessing-clean-unaids}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
   eval=T, 
  echo=T
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

### Introduction

This vignette covers how to access the clean and tidy annual UNAIDS Estimates from GitHub Releases. The MindTheGap package uses a cleaning function (`munge_unaids`) to process, munge, and tidy the UNAIDS HIV Estimates from AIDSInfo and `pull_unaids` to pull the cleaned data from Google Drive.

Let's start by loading the MindTheGap package.

```{r setup, message = FALSE}
library(mindthegap)
```

### Pull clean data from GitHub Releases

With `munge_unaids`, the developer is left with a clean and tidy dataset that includes a `pepfar` flag to denote whether a country is a PEPFAR country or not and an `estimate_flag` to indicate estimates that are an approximation (previously contained a "\<"). In order to make this data accessible outside of R and cut down on the processing time, we uploaded the clean UNAIDS data to the [package releases] (<https://github.com/USAID-OHA-SI/mindthegap/releases>) and wrote the function `pull_unaids` for users to easily access this data.

To use `pull_unaids`, there are 2 parameters:

1.  `data_type`: This parameter returns one of the 2 datasets: "HIV Estimates" or "HIV Test & Treat". Please note that if you want to filter to a specific indicator type (i.e. Percent or Integer), you will have to add that filter to your code `filter(indic_type == "Integer")`.

2.  `pepfar_only`: if TRUE, this parameter will return a dataset of only PEPFAR countries. If you do not specify this parameter, it will default to TRUE.

In order to access the data from GitHub releases, be sure to load the `glamr` package to run the `pull_unaids` function. 

```{r, message = FALSE}
remotes::install_github("USAID-OHA-SI/glamr")
library(glamr)

df_est <- pull_unaids(data_type = "HIV Estimates", pepfar_only = TRUE)
knitr::kable(head(df_est), format = "html")
```
