---
title: "Epidemic Control Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{epi_control_plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
   eval=F, 
  echo=T
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE, message = FALSE}
library(mindthegap)
library(knitr)
library(tidyverse)
library(kableExtra)
library(glamr)

load_secrets()
```

## Introduction

PEPFAR defines HIV epidemic control as the "point at which the total number of new HIV infections falls below the total number of deaths from all causes among HIV-infected individuals, with both declining." The integer indicators in the HIV Estimates data can be used to analyze and plot these trends. 

To assess progress towards this definition of epidemic control, we can use the `"HIV Estimates - Integer"` tab of the UNAIDS data. This vignette will walk through how to use the `epi_plot` function in the MindTheGap package to generate plots.

### Epidemic Control Plotting Function

First, let's take a look at the `epi_plot` function in action: to use this function, first load the `mindthegap` package and specify your PEPFAR country of interest in the `sel_cntry` parameter of the `epi_plot` function.

```{r, message = FALSE, warning=FALSE, fig.align="center"}

library(mindthegap)

epi_plot(sel_cntry = "Kenya")

```

In addition to looking at epidemic control at the country level, we can also see trends across all of PEPFAR by using `sel_cntry == "ALL PEPFAR"`.

```{r, message = FALSE, warning=FALSE, fig.align="center"}

library(mindthegap)

epi_plot(sel_cntry = "ALL PEPFAR")

```

This function makes it really easy to generate plots on your own for countries of interest. Let's peek under the hood and see how the function works to munge the data and create the plots.

### Reshaping epidemic control data

First, let's load all the other libraries that we need to generate these plots and load the `HIV Estimates - Integer` data.

```{r, message= FALSE}
library(glitr)
library(extrafont)
library(scales)
library(tidytext)
library(ggtext)
library(glue)

 df_epi <- mindthegap::pull_unaids(sheetname = "HIV Estimates - Integer", pepfar_only = TRUE)
  
```

- To create the epidemic control curves, you will need 2 indicators from the `HIV Estimates - Integer` tab:
    - **`New HIV Infections`**
    - **`AIDS Related Deaths`**
    
Let's first filter the data down to the 2 indicators that we need, and filter for all sexes and all ages.
    
```{r}
 #Filter down to the estimate and all ages, and the indicators you need
  df_epi_pepfar <- df_epi %>% 
    dplyr::filter(stat == "est",
           age == "all",
           sex == "all",
           indicator %in% c("AIDS Related Deaths", "New HIV Infections")) %>%
    dplyr::select(year, country, indicator, value) %>%
    dplyr::arrange(country, indicator, year) 
  
df_epi_pepfar
```

- Let's now perform some munging to easily identify some metrics that we care about for the PEPFAR definition of epidemic control.
    - **Declining deaths**: use `dplyr::lag()` function to create a `value = TRUE` if deaths are declining
    - **Infections below deaths**: `TRUE` wherever infections < deaths
    - **Ratio of infections / AIDS-related deaths**
    - **Epi_control**: `TRUE` if declining deaths and infections below deaths are both true

```{r}
  df_epi_pepfar <- df_epi_pepfar %>% 
    mutate(indicator = word(indicator, -1) %>% tolower) %>% #simplifies the indicator names
    pivot_wider(names_from = "indicator") %>% #pivots data wide into deaths and infections column
    group_by(country) %>% 
    mutate(declining_deaths = deaths - lag(deaths, order_by = year) <= 0) %>% #create a value that if true indicates declining deaths
    ungroup() %>% 
    mutate(infections_below_deaths = infections < deaths,
           ratio = infections / deaths,
           direction_streak = sequence(rle(declining_deaths)$lengths),
           epi_control = declining_deaths == TRUE & infections_below_deaths == TRUE) #epi control definition

df_epi_pepfar

```

Now that we have a workable dataframe, we can start to add color and style elements. Using `ifelse()` statements, we can add variables to indicate the colors of each indicator (drawing color inspiration from OHA's [Data Visualization Style Guide]((https://usaid-oha-si.github.io/styleguide/))). We can also add change the value for AIDS-related deaths to create the dual-axis.

```{r}
  #Add colors to indicators and flip axis
  df_epi_pepfar <- df_epi_pepfar %>% 
    pivot_longer(c(infections, deaths), names_to = "indicator") %>% 
    arrange(country, indicator, year) %>% 
    mutate(value_mod = ifelse(indicator == "deaths", -value, value),
           fill_color = ifelse(indicator == "deaths", old_rose, denim))
```

Let's filter our country down to Kenya now for a country-level epidemic control curve.

- We are also going to specify a couple more style elements to call on when we create our plot:
    - `val_lab`: Adding a value label for the most recent year of data. We can use the `scales::number()` function to standardize the number format.
    - `lab_pt`: Add a point at the most recent year's value

```{r}
#COUNTRY
    df_viz_cntry <- df_epi_pepfar %>%
      dplyr::filter(country %in% "Kenya") %>%
      dplyr::mutate(val_lab = dplyr::case_when(year == max(year) ~ scales::number(value, 1, scale = 1e-3, suffix = "k")),
             lab_pt = dplyr::case_when(year == max(year) ~ value_mod),
             country = factor(country, "Kenya"))
```
  
### Plotting the epidemic control curves

- Our dataframe is ready to go! Let's get started with the viz. We'll start by getting our framework down:
    - `ggplot()`: to define our aesthetics (x = `year`, y = `value_mod`, group by the indicator, and add fill and color as `fill_color`)
    - `geom_area()`: creates the area graph
    - `geom_line()`: creates the line border on the area graph
    - `geom_hline()`: adds a horizontal line at the x-axis

```{r}
df_viz_cntry %>%
      ggplot(aes(year, value_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() 
      # geom_point(aes(y = lab_pt), na.rm = TRUE,
      #            shape = 21, color = "white", size = 3) +
      # geom_text(aes(label = val_lab), na.rm = TRUE,
      #           hjust = -0.3,
      #           family = "Source Sans Pro Light") +
      # facet_wrap(~country, ncol = 2, scales = "free_y") +
      # scale_y_continuous(labels = ~ scales::label_number_si()(abs(.))) +
      # scale_x_continuous(breaks = seq(1990, 2025, 10)) +
      # scale_fill_identity(aesthetics = c("fill", "color")) +
      # labs(x = NULL, y = NULL) +
      # coord_cartesian(expand = T, clip = "off") +
      # glitr::si_style_ygrid() +
      # theme(axis.text.y = ggtext::element_markdown(),
      #       panel.spacing.x = unit(20, "pt"),
      #       panel.spacing.y = unit(0, "pt"))
```

- This is a good start. Let's continue to add style elements, adjust the scales, and add annotations
    - `geom_point()`: adds a point at `lab_pt` that we specified earlier
    - `geom_text()`: adds the annotation `val_lab` that we specified earlier
    - scale_y_continuous(): cleans up the y-axis with `scales::label_number_si()`
    - `scale_x_continuous()`: breaks the x-axis year markings into 10 year increments from 1990 to 2025.

```{r}
df_viz_cntry %>%
      ggplot(aes(year, value_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() +
       geom_point(aes(y = lab_pt), na.rm = TRUE,
                  shape = 21, color = "white", size = 3) +
      geom_text(aes(label = val_lab), na.rm = TRUE,
                hjust = -0.3,
                family = "Source Sans Pro Light") +
      scale_y_continuous(labels = ~ scales::label_number_si()(abs(.))) +
      scale_x_continuous(breaks = seq(1990, 2025, 10)) 
      #scale_fill_identity(aesthetics = c("fill", "color"))
      # labs(x = NULL, y = NULL) +
      # coord_cartesian(expand = T, clip = "off") +
      # glitr::si_style_ygrid() +
      # theme(axis.text.y = ggtext::element_markdown(),
      #       panel.spacing.x = unit(20, "pt"),
      #       panel.spacing.y = unit(0, "pt"))
```

We're almost there! Let's tackle the issue of color - even though we specified the color of the indicators in our aesthetics earlier, it still doesn't see to be registering our `fill_color.` To solve this issue, we have to use the function `scale_fill_identity()` to tell ggplot to use our pre-specified fill_color column as the color identity.

```{r}
df_viz_cntry %>%
      ggplot(aes(year, value_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() +
       geom_point(aes(y = lab_pt), na.rm = TRUE,
                  shape = 21, color = "white", size = 3) +
      geom_text(aes(label = val_lab), na.rm = TRUE,
                hjust = -0.3,
                family = "Source Sans Pro Light") +
      scale_y_continuous(labels = ~ scales::label_number_si()(abs(.))) +
      scale_x_continuous(breaks = seq(1990, 2025, 10))+
      scale_fill_identity(aesthetics = c("fill", "color"))
      # labs(x = NULL, y = NULL) +
      # coord_cartesian(expand = T, clip = "off") +
      # glitr::si_style_ygrid() +
      # theme(axis.text.y = ggtext::element_markdown(),
      #       panel.spacing.x = unit(20, "pt"),
      #       panel.spacing.y = unit(0, "pt"))
```

  
- Much better - we're almost there! Let's finish off our plot with some themes and styles.
    - `labs()`: By specifying x and y as `NULL`, we are removing the labels on the axis. Here is where you can add additional annotations such as titles, subtitles, captions, etc.
    - `coord_cartesian()`: changing cartesian plot and clipping preferences
    - `glitr::si_style_ygrid()`: uses standard SI style format from the `glitr` package
    - `theme()`: 

```{r}
df_viz_cntry %>%
      ggplot(aes(year, value_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() +
       geom_point(aes(y = lab_pt), na.rm = TRUE,
                  shape = 21, color = "white", size = 3) +
      geom_text(aes(label = val_lab), na.rm = TRUE,
                hjust = -0.3,
                family = "Source Sans Pro Light") +
      scale_y_continuous(labels = ~ scales::label_number_si()(abs(.))) +
      scale_x_continuous(breaks = seq(1990, 2025, 10))+
      scale_fill_identity(aesthetics = c("fill", "color")) +
      facet_wrap(~country, ncol = 2, scales = "free_y") +
      labs(x = NULL, y = NULL) +
      coord_cartesian(expand = T, clip = "off") +
      glitr::si_style_ygrid() +
      theme(axis.text.y = ggtext::element_markdown(),
            panel.spacing.x = unit(20, "pt"),
            panel.spacing.y = unit(0, "pt"))
```
