---
title: "Epidemic Control Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{epi_control_plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
   eval=T, 
  echo=T
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE, message = FALSE}
library(mindthegap)
library(knitr)
install.packages("tidyverse")
library(tidyverse)
library(kableExtra)
library(glamr)
remotes::install_github("USAID-OHA-SI/glitr")
library(glitr)
install.packages("ggtext")
library(ggtext)

```

## Introduction

PEPFAR defines HIV epidemic control as the "point at which the total number of new HIV infections falls below the total number of deaths from all causes among HIV-infected individuals, with both declining." The integer indicators in the HIV Estimates data can be used to analyze and plot these trends.

To assess progress towards this definition of epidemic control, we can use the `"HIV Estimates"` and `"epicontrol"` tab of the UNAIDS data. This vignette will walk through how to use the `epi_plot` function in the MindTheGap package to generate plots.

### Epidemic Control Plotting Function

First, let's take a look at the `epi_plot` function in action: to use this function, first load the `mindthegap` package and specify your PEPFAR country/countries of interest in the `sel_cntry` parameter of the `epi_plot` function.

```{r, message = FALSE, warning=FALSE, fig.show = "hold", out.width="50%"}

#Load library 
library(mindthegap)

epi_plot(sel_cntry = "Kenya")
epi_plot(sel_cntry = c("Kenya","Mozambique", "Tanzania"))

```

In addition to looking at epidemic control at the country level, we can also see trends across all of PEPFAR by just using the `epi_plot()` function, as "ALL PEPFAR" is the default country in the `sel_cntry` parameter.

```{r, message = FALSE, warning=FALSE, fig.align="center"}

library(mindthegap)

epi_plot()

```

This function makes it really easy to generate plots on your own for countries of interest. Let's peek under the hood and see how the function works to munge the data and create the plots.

### Reshaping epidemic control data

First, let's load all the other libraries that we need to generate these plots and load the `HIV Estimates` data.

```{r, message= FALSE}
library(glitr)
library(extrafont)
library(scales)
library(tidytext)
library(ggtext)
library(glue)
library(magrittr)
library(ggplot2)

 df_epi <- mindthegap::pull_unaids(data_type = "HIV Estimates", pepfar_only = TRUE)
 
```

-   To create the epidemic control curves, you will need 2 indicators from the `HIV Estimates` tab:
    -   **`Number New HIV Infections`**
    -   **`Total deaths to HIV Population`**

Let's first filter the 2 datasets down to the indicators that we need, and filter for all sexes and all ages.

```{r}
#Filter down to the estimate and all ages, and the indicators you need

  #Pull indicators 
 df_epi_pepfar <- df_epi %>%
 dplyr::filter(age == "All", sex == "All",
        indicator %in% c("Total deaths to HIV Population", "Number New HIV Infections")) %>% #grab indicators 
      dplyr::select(year, country,indicator, estimate) %>% 
      dplyr::arrange(country, indicator, year) #order rows by these variables 

df_epi_pepfar

```

-   Let's now merge the datasets and perform some munging to easily identify some metrics that we care about for the PEPFAR definition of epidemic control.
    -   **Declining deaths**: use `dplyr::lag()` function to create a `value = TRUE` if deaths are declining
    -   **Infections below deaths**: `TRUE` wherever infections \< deaths
    -   **Ratio of infections / deaths**
    -   **Epi_control**: `TRUE` if declining deaths and infections below deaths are both true

```{r}
#Create the merged data frame  
    
    #Perform necessary munging
    df_epi_pepfar <- df_epi_pepfar %>% 
      tidyr::pivot_wider(names_from = "indicator", 
                         values_from = "estimate", #pivots data wide into infections column
                         names_glue = "{indicator %>% stringr::str_extract_all('deaths|Infections') %>% tolower}") 

    #Add in ALL PEPFAR data
    df_epi_pepfar <-
      df_epi_pepfar %>%
      dplyr::bind_rows(df_epi_pepfar %>%
                  dplyr::mutate(country = "All PEPFAR") %>%
                  dplyr::group_by(country, year) %>%
                  dplyr::summarise(across(where(is.numeric), \(x) sum(x,na.rm = TRUE)), .groups = "drop")) #sums PEPFAR country estimates 
    
   #Create epi control flag
    df_epi_pepfar <-
      df_epi_pepfar %>%
      dplyr::mutate(declining_deaths = deaths - dplyr::lag(deaths, order_by = year) <= 0, by = c(country)) %>% #TRUE/FALSE declining 
      dplyr::mutate(infections_below_deaths = infections < deaths,
             ratio = infections / deaths,
             direction_streak = sequence(rle(declining_deaths)$lengths),
             epi_control = declining_deaths == TRUE & infections_below_deaths == TRUE) #epi control definition 
      
   
 
df_epi_pepfar
```

Now that we have a workable data frame, we can start to add color and style elements. Using `ifelse()` statements, we can create the `fill_color` variable to indicate the colors of each indicator (drawing color inspiration from OHA's [Data Visualization Style Guide]((https://usaid-oha-si.github.io/styleguide/))). We can also create the `val_mod` variable to add negative values for total deaths in order to create the dual-axis.

```{r}
  #Add colors to indicators and flip axis
  df_epi_pepfar <- df_epi_pepfar %>% 
    tidyr::pivot_longer(c(infections, deaths), names_to = "indicator") %>% #put back indicators in column
    dplyr::arrange(country, indicator, year) %>%
    dplyr::mutate(val_mod = ifelse(indicator == "deaths", -value, value), #create dual-axis
                  fill_color = ifelse(indicator == "deaths", glitr::old_rose, glitr::denim)) #add colors to indicate flip axis

df_epi_pepfar
```

Let's filter our country down to Kenya now for a country-level epidemic control curve.

-   We are also going to specify a couple more style elements to call on when we create our plot:
    -   `val_lab`: adding a value label for the most recent year of data. We can use the `scales::number()` function to standardize the number format.
    -   `max_plot_pt` & `min_plot_pt`: define a max and min for the dual-axis
    -   `lab_pt`: add a point at the most recent year's value
    -   `new_hiv_label` & `tot_death_label`: create labels for the indicators and place them at the max/min y-axis point

```{r}
#COUNTRY
    df_viz_cntry <- df_epi_pepfar %>%
      dplyr::filter(country %in% "Kenya") %>%
      dplyr::mutate(val_lab = dplyr::case_when(year == max(year) ~ 
                                                 scales::number(value, 1, scale = 1e-3, suffix = "k")), #standardize number format
                    max_plot_pt = max(value),
                    min_plot_pt = min(val_mod),
             lab_pt = dplyr::case_when(year == max(year) ~ val_mod),
             indicator = ifelse(indicator == "deaths", "Total Deaths to HIV Population", "New HIV Infections"), #creating labels
             new_hiv_label = dplyr::case_when(value == max_plot_pt ~ indicator),  #assigning label location to min/max
             tot_death_label = dplyr::case_when(val_mod == min_plot_pt ~ indicator)) %>%
      dplyr::mutate(cntry_order = max(value, na.rm = T), .by = country) %>%
      dplyr::mutate(country = forcats::fct_reorder(country, cntry_order, .desc = T))
    
```

### Plotting the epidemic control curves

-   Our data frame is ready to go! Let's get started with the viz. We'll start by getting our framework down:
    -   `ggplot()`: to define our aesthetics (x = `year`, y = `val_mod`, group by the indicator, and add fill and color as `fill_color`)
    -   `geom_blank()`: defines a max point on the y-axis
    -   `geom_area()`: creates the area graph
    -   `geom_line()`: creates the line border on the area graph
    -   `geom_hline()`: adds a horizontal line at the x-axis

```{r}
df_viz_cntry %>%
      ggplot(aes(year, val_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_blank(aes(y = max_plot_pt)) + #sets max y-axis above
      geom_blank(aes(y = -max_plot_pt)) + #sets max y-axis below 
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() 

```

-   This is a good start. Let's continue to add style elements, adjust the scales, and add annotations
    -   `geom_point()`: adds a point at `lab_pt` that we specified earlier
    -   `geom_text()`: adds the annotation `val_lab` that we specified earlier
    -   `scale_y_continuous()`: cleans up the y-axis with `scales::label_number()`and `expand`
    -   `scale_x_continuous()`: takes the max and min year and breaks the x-axis year markings into 5 year increments.

```{r}
df_viz_cntry %>%
      ggplot(aes(year, val_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_blank(aes(y = max_plot_pt)) + #sets max y-axis above
      geom_blank(aes(y = -max_plot_pt)) + #sets max y-axis below 
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line()+
      geom_point(aes(y = lab_pt), na.rm = TRUE, shape = 21, color = "white", size = 3) +
      geom_text(aes(label = val_lab), na.rm = TRUE,
                hjust = -0.3,
                family = "Source Sans Pro Light") + #value label text
      scale_y_continuous(labels = ~ (scales::label_number(scale_cut = scales::cut_short_scale())(abs(.))),
                                  expand = c(0, 0)) +
      scale_x_continuous(breaks = seq(min(df_epi_pepfar$year), max(df_epi_pepfar$year),5))#automatic x-axis min/max

```

We're almost there! Let's tackle the issue of color - even though we specified the color of the indicators in our aesthetics earlier, it still doesn't seem to be registering our `fill_color.` To solve this issue, we have to use the function `scale_fill_identity()` to tell ggplot to use our pre-specified fill_color column as the color identity.

```{r}
df_viz_cntry %>%
      ggplot(aes(year, val_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() +
       geom_point(aes(y = lab_pt), na.rm = TRUE,
                  shape = 21, color = "white", size = 3) +
      geom_text(aes(label = val_lab), na.rm = TRUE,
                hjust = -0.3,
                family = "Source Sans Pro Light") +
       scale_y_continuous(labels = ~ (scales::label_number(scale_cut = scales::cut_short_scale())(abs(.))),
                                  expand = c(0, 0)) +
      scale_x_continuous(breaks = seq(min(df_epi_pepfar$year), max(df_epi_pepfar$year),5))+    
      scale_fill_identity(aesthetics = c("fill", "color"))

```

-   Much better - we're almost there! Let's finish off our plot with some themes and styles.
    -   `geom_text`: adds the annotation for the `new_hiv_label` and `tot_death_label` we created earlier
    -   `labs()`: By specifying x and y as `NULL`, we are removing the labels on the axis. Here is where you can add additional annotations such as titles, subtitles, captions, etc.
    -   `coord_cartesian()`: changing cartesian plot and clipping preferences
    -   `glitr::si_style_ygrid()`: uses standard SI style format from the `glitr` package and `facet_space` adjusts the spacing of the y-axis grid
    -   `theme()`: allows you to customize components of the plot with `axis.text.y` focusing on the y-axis text

```{r}
df_viz_cntry %>%
      ggplot(aes(year, val_mod, group = indicator, fill = fill_color, color = fill_color)) +
      geom_area(alpha = .25) +
      geom_hline(yintercept = 0, color = glitr::grey80k) +
      geom_line() +
       geom_point(aes(y = lab_pt), na.rm = TRUE,
                  shape = 21, color = "white", size = 3) +
      geom_text(aes(label = val_lab), na.rm = TRUE,
                hjust = -0.3,
                family = "Source Sans Pro Light") +
      scale_y_continuous(labels = ~ (scales::label_number(scale_cut = scales::cut_short_scale())(abs(.))),
                                  expand = c(0, 0)) +
      scale_x_continuous(breaks = seq(min(df_epi_pepfar$year), max(df_epi_pepfar$year),5))+ 
      scale_fill_identity(aesthetics = c("fill", "color")) +
        ggplot2::facet_wrap(~country) + #small multiples of countries
  geom_text(aes(label = new_hiv_label, x = 2005, y = (max_plot_pt)), na.rm = TRUE,
                               hjust = -0.3, family = "Source Sans Pro Light") +
  geom_text(aes(label = tot_death_label, x = 2005, y = (min_plot_pt)), na.rm = TRUE,
                               hjust = -0.3, family = "Source Sans Pro Light") +
      labs(x = NULL, y = NULL) + coord_cartesian(expand = T, clip = "off") +
      glitr::si_style_ygrid(facet_space = 0.75) + #adjusted y-axis grid spacing 
      theme(axis.text.y = ggtext::element_markdown()) +
      labs(caption = "Source: UNAIDS Data 2022 Release")
```
