---
title: "Data Dictionary & Use Cases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{unaids-data-dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE, message = FALSE}
library(mindthegap)
library(knitr)
library(tidyverse)
library(kableExtra)
library(glamr)
load_secrets()
```


## Introduction

This vignette serves as a data dictionary for the UNAIDS 2020 Estimates Data. It will cover the data structure and variables, as well as provide an overview of the indicators across the HIV Estimates tab and the Test & Treat tab. Finally, it will touch briefly on common use cases and which data tab to access for those analytics. 

## UNAIDS Clean Data Structure

Let's first take a look at the data structure, using the `HIV Estimates - Percent` tab as an example. 

```{r, message = FALSE}
df_pct <- pull_unaids(sheetname = "HIV Estimates - Percent", pepfar_only = TRUE)
knitr::kable(head(df_pct, n = 6L), format = "html")

```

Within each of the UNAIDS datasets, there are disaggregates by `age` and `sex.` There is also a variable called `stat`, which indicates whether the value is the actual estimate, or the lower or upper bound. Let's take a look at the disaggregates in the `"HIV Estimates - Percent"` tab. 

```{r, message = FALSE, echo = FALSE}
count_tbl <- df_pct %>% 
  count(indicator, age, sex, stat)

knitr::kable(count_tbl,
             format = "html",
             caption = "Age/Sex/Stat Count Table from the `HIV Estimates - Percent` tab") %>% 
    kable_styling(bootstrap_options = c("hover"))
```

As you can see, there are multiple different disaggregates, especially across indicators. For instance, if you were looking at HIV Incidence, there are age disaggregates for ages 15-49 and all ages, but no disaggregates for sex. If you were looking at Prevalence, there are sex disaggregates for young adults (15-24), but no sex disaggregates for the age group 15-49.

As such, it is important to filter for the age/sex/stat disaggregates that are you interested in before diving further into your analytics.

## HIV Estimates Indicators and Use Cases

The HIV estimates data covers epidemiological indicators, such as HIV prevalence/incidence, new infections, AIDS-related death, etc. There are 2 percent indicators (HIV Prevalence and HIV Incidence) and 4 integer indicators (PLHIV, AIDS-related deaths, PMTCT, and New HIV Infections).

To pull the percent HIV Estimates indicators for PEPFAR countries, use `pull_unaids(sheetname = "HIV Estimates - Percent, pepfar_only = TRUE)`.

To pull the integer HIV Estimates indicators, use `pull_unaids(sheetname = "HIV Estimates - Integer, pepfar_only = TRUE)`.

```{r, echo = FALSE, message = FALSE}
hiv_est_info <- tribble(
    ~ indicator, ~ description,
  "`Prevalence`",   "HIV Prevalence",
  "`Incidence`", "HIV Incidence (per 1,000 uninfected population)",
  "`PLHIV`", "Estimated people living with HIV",
  "`AIDS Related Deaths`", "AIDS Related Deaths",
  "`PMTCT`", "Pregnant women needing antiretrovirals for preventing mother-to-child transmission",
  "`New HIV Infections`", "New HIV Infections",
)

kable(
  hiv_est_info,
  format = "html",
  col.names = c("", "")
) %>%
  group_rows(index = c("Percent Indicators" = 2, "Integer Indicators" = 4)) %>%
  kable_styling(bootstrap_options = c("hover"))
```

### Use Case #1: Epidemic Control Curves

PEPFAR defines HIV epidemic control as the "point at which the total number of new HIV infections falls below the total number of deaths from all causes among HIV-infected individuals, with both declining." The integer indicators in the HIV Estimates data can be used to analyze and plot these trends. 

To assess progress towards this definition of epidemic control, we will use the `"HIV Estimates - Integer"` data:

```{r, message = FALSE}
df_est <- pull_unaids(sheetname = "HIV Estimates - Integer", pepfar_only = TRUE)
```

```{r, echo = FALSE, message = FALSE}
hiv_est_info <- tribble(
    ~ indicator, ~ description,
  "`PLHIV`", "Estimated people living with HIV",
  "`AIDS Related Deaths`", "AIDS Related Deaths",
  "`PMTCT`", "Pregnant women needing antiretrovirals for preventing mother-to-child transmission",
  "`New HIV Infections`", "New HIV Infections",
)

kable(
  hiv_est_info,
  format = "html",
  col.names = c("", "")
) %>%
  group_rows(index = c("Integer Indicators" = 4)) %>%
  kable_styling(bootstrap_options = c("hover"))
```

- To create the epidemic control curves, you will need 2 indicators from the `HIV Estimates - Integer` tab:
    - **`New HIV Infections`**
    - **`AIDS Related Deaths`**

For more information on how to create these plots, see the article `Epidemic Control Plots.`


### Use Case #2: Incidence/Prevalence Curves

To plot HIV Incidence and Prevalence curves with confidence intervals, we will use the `"HIV Estimates - Percent"` data:

```{r, message = FALSE}
df_est <- pull_unaids(sheetname = "HIV Estimates - Percent", pepfar_only = TRUE)
```

```{r, echo = FALSE, message = FALSE}
hiv_est_info <- tribble(
    ~ indicator, ~ description,
  "`Prevalence`",   "HIV Prevalence",
  "`Incidence`", "HIV Incidence (per 1,000 uninfected population)",
)

kable(
  hiv_est_info,
  format = "html",
  col.names = c("", "")
) %>%
  group_rows(index = c("Percent Indicators" = 2)) %>%
  kable_styling(bootstrap_options = c("hover"))
```

- To create the incidence/prevalence curves, you will need 2 indicators from the `HIV Estimates - Percent` tab: 
    - **`Prevalence`**
    - **`Incidence`**

## Test & Treat Indicators & Use Cases

The Test & Treat Data covers indicators in the clinical cascade, that can be used to track progress to the UNAIDS 90-90-90 targets. There are 6 percent indicators and 5 integer indicators.

To pull the percent Test & Treat indicators for PEPFAR countries, use `pull_unaids(sheetname = "Test & Treat - Percent, pepfar_only = TRUE)`.

To pull the integer Test & Treat indicators, use `pull_unaids(sheetname = "Test & Treat - Integer, pepfar_only = TRUE)`.

```{r,echo = FALSE}
hiv_est_info <- tribble(
    ~ indicator, ~ description,
  "`KNOWN_STATUS`",   "**Among PLHIV**, Percent who known their HIV status",
  "`PLHIV_ON_ART`", "**Among PLHIV**, Percent on ART",
  "`KNOWN_STATUS_ON_ART`", "**Among people who know their HIV status**, Percent of ART",
  "`VLS`", "**Among PLHIV**, Percent with Suppressed Viral Load",
  "`ON_ART_VLS`", "**Among people on ART**, Percent with Suppressed Viral Load",
  "`PMTCT_ON_ART`", "Estimated Percentage of Women Living with HIV on ART for preventing mother-to-child transmission",
  
  "`KNOWN_STATUS`", "Number who know their HIV status",
  "`PLHIV_ON_ART`", "Number on ART",
  "`VLS`", "Number with Suppressed Viral Load",
  "`PMTCT`", "Number of Pregnant women needing antiretrovirals for preventing mother-to-child transmission",
  "`PMTCT_ON_ART`", "Estimated Number of Women Living with HIV on ART for preventing mother-to-child transmission",
)

kable(
  hiv_est_info,
  format = "html",
  col.names = c("", "")
) %>%
  group_rows(index = c("Percent Indicators" = 6, "Integer Indicators" = 4)) %>%
  kable_styling(bootstrap_options = c("hover"))
```

### Use Case #3: Progress to 90-90-90 Targets

The percent indicators in the Test & Treat data can be used to assess progress to the UNAIDS 90-90-90 targets that were set to be achieved in 2020, where 90% of PLHIV know their status, 90% of those who know their status are accessing treatment, and 90% of those who are accessing treatment are virally suppressed.

To assess progress to 90-90-90 targets, we will use the `"Test & Treat - Percent"` data:

```{r, message = FALSE}
df_tt <- pull_unaids(sheetname = "Test & Treat - Percent", pepfar_only = TRUE)
```


```{r, echo = FALSE}
hiv_est_info <- tribble(
    ~ indicator, ~ description,
  "`KNOWN_STATUS`",   "**Among PLHIV**, Percent who known their HIV status",
  "`PLHIV_ON_ART`", "**Among PLHIV**, Percent on ART",
  "`KNOWN_STATUS_ON_ART`", "**Among people who know their HIV status**, Percent of ART",
  "`VLS`", "**Among PLHIV**, Percent with Suppressed Viral Load",
  "`ON_ART_VLS`", "**Among people on ART**, Percent with Suppressed Viral Load",
  "`PMTCT_ON_ART`", "Estimated Percentage of Women Living with HIV on ART for preventing mother-to-child transmission"
)

kable(
  hiv_est_info,
  format = "html",
  col.names = c("", "")
) %>%
  group_rows(index = c("Percent Indicators" = 6)) %>%
  kable_styling(bootstrap_options = c("hover"))
```

As you may notice, there are multiple test & treat percent indicators for the clinical cascade. This is because there are 2 types of metrics that can be used to calculate progress to 90-90-90 targets - the **relative base/conditional denominator** or the **PLHIV base/denominator**.

- If you are using the **relative base** to track 90-90-90 progress, use the following indicators from the `"Test & Treat - Percent"` data tab.
    - **`KNOWN_STATUS`**: Percent of PLHIV who known their HIV status
    - **`KNOWN_STATUS_ON_ART`**: Among those who know their status, percent on ART
    - **`ON_ART_VLS`**: Among those who are on ART, percent virally suppressed
    
- If you are using the **PLHIV base**, use the following indicators from the "Test & Treat - Percent" data tab.
    - **`KNOWN_STATUS`**: Percent of PLHIV who known their HIV status
    - **`PLHIV_ON_ART`**: Among PLHVIV, percent on ART
    - **`VLS`**: Among PLHIV, percent virally suppressed
    
For more information about the distinction between relative/PLHIV base and how to assess 90-90-90 progress, see the article `90-90-90 Target Progress: Relative Base vs. PLHIV Base`.

    
